Highly Available Transactions: Virtues and Limitations

### 2013-05-01 writeup/conclusion.tex
pass, plus spelling and grammar

\section{Conclusions and Future Work}
The current state of database software offers uncomfortable and
unnecessary choices between availability and transactional semantics.
Through our analysis and experiments, we have demonstrated how goals
of high availability will remain a critical aspect of many future data
storage systems. Accordingly, we expose a broad design space of Highly
Available Transactions (HATs), which can offer the key benefits of
highly available distributed systems---``always on'' operation during
partitions and low-latency operations---while also providing a family
of transactional isolation levels and replica semantics that have been
adopted in practice.  In taxonomizing this design space, we show that
many transactional guarantees are achievable with high availability,
while a few, like preventing Lost Update and Write Skew, are not. This
offers a useful connection between storage semantics studied in the
database and distributed systems communities.

>>>>

\section{Conclusions and Future Work}
\label{sec:conclusion}

The current state of database software offers uncomfortable and
unnecessary choices between availability and transactional semantics.
Through our analysis and experiments, we have demonstrated how goals
of high availability will remain a critical aspect of many future data
storage systems. Accordingly, we expose a broad design space of Highly
Available Transactions (HATs), which can offer the key benefits of
highly available distributed systems---``always on'' operation during
partitions and low-latency operations---while also providing a family
of transactional isolation levels and replica semantics that have been
adopted in practice.  In taxonomizing this design space, we show that
many transactional guarantees are achievable with high availability,
while a few, like preventing Lost Update and Write Skew, are not. This
offers a useful connection between storage semantics studied in the
database and distributed systems communities.

### 2013-05-01 writeup/evaluation.tex
pass, plus spelling and grammar

\begin{myenumerate}\vspace{-.5em}
\item We revisit traditional database concurrency control with a focus
  on coordination costs and on high availability.
\item We examine the properties required by a realistic OLTP
  application based on the TPC-C benchmark.
\item We perform a brief experimental evaluation of HAT versus non-HAT
  properties on public cloud infrastructure.
\end{myenumerate}

\subsection{Existing Algorithms}

@@

\vspace{.5em}\noindent\textbf{Serializability} To establish a serial
order on transactions, algorithms for achieving serializability of
general-purpose read-write transactions in a distributed
setting~\cite{bernstein-book, davidson-survey, wisemann-survey}
require at least one round trip time (RTT) before committing. As an
example, traditional two-phase locking for a transaction of length $T$
may require $T$ \texttt{lock} operations and will require at least one
\texttt{lock} and one \texttt{unlock} operation.  In a distributed
environment, each of these operations requires coordination, either
with other database servers or with a lock service. If this
coordination mechanism is unavailable, transactions cannot safely
commit. Similarly, optimistic concurrency control requires
coordinating via a validation step, while deterministic transaction
scheduling~\cite{deterministic-scheduling} requires a
scheduler. Serializability under multi-version concurrency control
requires checking for update conflicts. All told, the reliance on a
globally agreed total order necessitates a minimum of one round-trip
to a designated master or coordination service for each of these
classic algorithms.  The cost of round trips will be determined by the
deployment environment, as we saw in Section~\ref{sec:motivation}; we
will demonstrate this cost on public cloud infrastructure in
Section~\ref{sec:prototype}.

\vspace{.5em}\noindent\textbf{Non-serializability} Most existing
distributed implementations of weak isolation are not highly
available. Lock-based proposals such as those used to provide weak
isolation in Gray's original proposal~\cite{gray-isolation} do not
degrade gracefully in the presence of partial failures. (Note,
however, that lock-based protocols \textit{do} offer the benefit of
recency guarantees.) While multi-versioned storage systems allow for a
variety of transactional guarantees, few offer traditional weak
isolation (e.g., non-``tentative update'' schemes) in this context.
The MDCC~\cite{mdcc} protocol offers Read Committed isolation with
Lost Update avoidance but is similarly unavailable due to its reliance
on preventing write conflicts. Chan and Gray's read-only transactions
have item-cut isolation with causal consistency and TA (session
PL-2L~\cite{adya}) but are unavailable in the presence of coordinator
failure and assume serializable update transactions~\cite{readonly};
this is similar to read-only and write-only transactions more recently
proposed by Eiger~\cite{eiger}. Causal Serializability offers a
similar model (with unavailable implementation): causal consistency
with a variant of Read Uncommitted between transactions that write to
the same data item~\cite{raynal-causal}.  Brantner's S3
database~\cite{kraska-s3} and Bayou~\cite{sessionguarantees} can all
provide variants of session PL-2L with high availability, but none
provide this HAT functionality without substantial
modification. Swift~\cite{swift} and bolt-on causal
consistency~\cite{bolton} are closest to providing maximum sticky HAT
semantics. As we have seen, it is possible to implement many
guarantees weaker than serializability---including guarantees that are
achievable with high availability---and still not achieve high
availability.

@@

\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} Recent
work on the CALM Theorem~\cite{calm} and Commutative and Replicated
Data Types~\cite{crdt} demonstrates that, if updates logically
commute, then they can often be safely performed in different orders
at different replicas. Accordingly, as long as all writes are
delivered to all replicas, then a system executing monotonic logic
with commutative operators may not suffer from application-level
consistency anomalies as a result of Lost Update or Write Skew
anomalies. However, applications with non-monotonic state mutation
will not, in general, be able to maintain application-level
consistency constraints with HATs alone---in particular, applications
requiring bounded update visibility latency should opt for
unavailability.

@@

To further understand the performance implications of HAT guarantees
in a real-world environment, we implemented a HAT database
prototype. Across a range of deployments on public cloud
infrastructures, we found that, as Section~\ref{sec:latency}'s
measurements suggested, ``strongly consistent'' algorithms incur
substantial latency penalties: over WAN, 10 to 100 times higher than
their HAT counterparts. In light of recent algorithms for providing
causally consistent get transactions~\cite{bolton, cops}, we focus our
HAT analysis on TA, which we believe to be the most novel of our
algorithms. We find that HAT properties---while more
resource-intensive than basic eventual consistency---scale well to
many servers.

\vspace{.5em}\noindent\textit{Implementation.} Our prototype database
is a partially replicated (hash-based, partitioned) key-value backed
by LevelDB and implemented in Java using Apache Thrift. It currently
supports eventual consistency (last-writer-wins RU with standard
all-to-all anti-entropy between replicas) and HAT RC and TA guarantees
(as described in Section~\ref{sec:hats}). It also supports non-HAT
operation whereby all operations for a given key are routed to a
(randomly) designated \texttt{master} replica for each key (i.e.,
PNUTS~\cite{pnuts}) as well as distributed two-phase locking. Servers
are durable: they synchronously write to LevelDB before responding to
client requests, while pending writes in TA are sent to a write-ahead
log, which is synchronously flushed to disk.

\vspace{.5em}\noindent\textit{Configuration.} We deploy the database
in \textit{clusters}---disjoint sets of database servers that each
contain a single, fully replicated copy of the data---typically across
datacenters, sticking all clients within a datacenter to their
respective cluster (trivially providing read-your-writes and monotonic
reads guarantees). By default, we use 5 Amazon EC2 \texttt{m1.xlarge}
instances as servers in each cluster. For our workload, we link our
client library to the YCSB benchmark~\cite{ycsb}, which is well-suited
to LevelDB's key-value schema, grouping every eight YCSB operations
from the default workload (50\% reads, 50\% writes) to form a
transaction. We increase the number of keys in the workload from the
default $1,000$ to $100,000$ with uniform random key access, keeping
the default value size of $1KB$, and running YCSB for 180 seconds per
configuration.

@@

\vspace{.5em}\noindent\textit{Geo-replication.} We first deploy the
database prototype across an increasing number of
datacenters. Figure~\ref{fig:wan-exp}A shows that, when operating two
clusters within a single datacenter, employing a single master for
each data item results in approximately half the throughput and double
the latency of an eventually consistent configuration: the HAT models
are able to utilize double the number of servers. RC---essentially
eventual consistency with buffering---is almost identical to eventual
consistency, while TA---which incurs two writes for every client-side
write (i.e., pending writes are sent to a WAL then subsequently moved
into LevelDB once stable)---achieves ~75\% of the
throughput. Latency increases linearly with the number of YCSB clients
due to contention within LevelDB.

 In contrast, when the two clusters are deployed across the
 continental United States (Figure~\ref{fig:wan-exp}B), the average
 latency of mastered operation increases to $300$ms (a $278$--$4257\%$
 latency increase; average $37$ms latency per operation). For the same
 number of YCSB client threads, mastered operation has substantially
 lower throughput than the HAT configurations. Increasing the number
 of YCSB clients \textit{does} increase the throughput of mastered
 operation, but our Thrift-based server-side connection processing did
 not gracefully handle more than several thousand concurrent
 connections. In contrast, across two datacenters, the performance of
 eventual, RC, and TA is near-identical to a single-datacenter
 deployment.

When we deployed five clusters (as opposed to two, as before) across
the five EC2 datacenters with lowest communication cost
(Figure~\ref{fig:wan-exp}C), the trend continues: mastered latency
increases to nearly $800$ms per transaction. As an attempt at reducing
this overhead, we implemented and benchmarked a variant of
quorum-based replication as in Dynamo~\cite{dynamo}, where clients
sent requests to all replicas, which completed as soon as a majority
of servers responded (guaranteeing regular
semantics~\cite{herlihy-art}); this strategy (not pictured) did not
substantially improve performance due to the network topology and
because worst-case server load was unaffected. With five clusters,
TA's relative throughput decreased: every YCSB \texttt{put} operation
resulted in four \texttt{put} operations on remote replicas and,
accordingly, the cost of anti-entropy increased (e.g., each server
processed four replica's anti-entropy as opposed to one before,
reducing the opportunity for batching and decreasing available
resources for for incoming client requests). This in turn increased
garbage collection activity and, more importantly, IOPS when compared
to Eventual and RC, causing TA throughput to peak at around half of
Eventual. With an in-memory persistence (i.e., no LevelDB or WAL), TA
throughput was within 20\% of eventual.

We have intentionally omitted performance data for two-phase
locking. Mastered configuration performed \textit{far} better than our
textbook implementation, which, in addition to requiring a WAN
round-trip per transaction, also incurred substantial overheads due to
mutual exclusion via locking. We expect that, while techniques like
those recently proposed in Calvin~\cite{calvin} can reduce the
overhead of serializable transactions by avoiding locking, our
master-based implementation and the data from
Section~\ref{sec:latency} are reasonable lower bounds on latency.

@@

\vspace{.5em}\noindent\textit{Transaction length.} As shown in
Figure~\ref{fig:txlen} (clusters in Virginia and Oregon), throughput
of eventual consistency, RC, and mastered operation are unaffected by
transaction length. In contrast, TA's throughput decreases linearly
with increased transaction length: with $1$ operation per transaction,
TA throughput is within 18\% of eventual ($34$ bytes overhead), and
with $128$ operations per transaction, TA throughput is within $60\%$
($1898$ bytes overhead). This reflects our TA algorithm's metadata
requirements, which are proportional to transaction length and consume
IOPS and network bandwidth.

\vspace{.5em}\noindent\textit{Read proportion.} Our default (equal)
proportion of reads and writes is fairly pessimistic: for example,
Facebook reports $99.8\%$ reads for their workload~\cite{eiger}. As
shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
with all reads, TA is within $4.8\%$ of eventual; with all writes, TA
is within $33\%$, and the throughput of eventual consistency decreases
by $288.8\%$ compared to all reads. At $99.8\%$ reads, TA incurs a $7\%$ overhead ($5.8\%$ for
in-memory storage).

\vspace{.5em}\noindent\textit{Scale-out.} One of the key benefits of
our HAT algorithms is that they are shared-nothing, meaning they
should not compromise scalability. Figure~\ref{fig:scaleout} shows
that varying the number of servers across two clusters in Virginia and
Oregon (with $15$ YCSB clients per server) results in linear scale-out
for both eventual, RC, and TA. RC and eventual scale perfectly with an
approximately $5$x throughput increase moving from $5$ to $25$ servers
per cluster. For the same configuration, TA scales by $3.8$x,
achieving over $260,000$ operations per second. We believe that TA's
performance is due to resource contention---with a memory-backed
database (instead of LevelDB), TA scales by $4.25$x (not shown)---and
TA-related performance heterogeneity across servers (Calvin's authors
report similar heterogeneity on EC2~\cite{calvin}). Nevertheless, this
linear scalability is not typical of traditional database system
designs~\cite{gray-isolation}.

\vspace{.5em}\noindent\textit{Summary.} Our experimental prototype
confirms our earlier analytical intuitions. HAT systems can provide
useful semantics without substantial performance penalties. In
particular, our TA algorithm can achieve throughput competitive with
eventual consistency at the expense of increased disk and network
utilization. Perhaps more importantly, all HAT algorithms circumvent
high WAN latencies inevitable with non-HAT implementations. Our results
highlight Deutcsh's observation that ignoring factors such as latency
can ``cause big trouble and painful learning
experiences''~\cite{fallacies-deutsch}---in a single-site context,
paying the cost of coordination may be tenable, but, especially as
services are geo-replicated, costs increase.

>>>>

\begin{myenumerate}\vspace{-.5em}
\item We revisit traditional database concurrency control with a focus
  on coordination costs and on high availability.
\item We examine the properties required by an OLTP application based
  on the TPC-C benchmark.
\item We perform a brief experimental evaluation of HAT versus non-HAT
  properties on public cloud infrastructure.
\end{myenumerate}

\subsection{Existing Algorithms}
\label{sec:eval-existing}

@@

\vspace{.5em}\noindent\textbf{Serializability} To establish a serial
order on transactions, algorithms for achieving serializability of
general-purpose read-write transactions in a distributed
setting~\cite{bernstein-book, davidson-survey, wisemann-survey}
require at least one RTT before committing. As an
example, traditional two-phase locking for a transaction of length $T$
may require $T$ \texttt{lock} operations and will require at least one
\texttt{lock} and one \texttt{unlock} operation.  In a distributed
environment, each of these lock operations requires coordination,
either with other database servers or with a lock service. If this
coordination mechanism is unavailable, transactions cannot safely
commit. Similarly, optimistic concurrency control requires
coordinating via a validation step, while deterministic transaction
scheduling~\cite{deterministic-scheduling} requires a
scheduler. Serializability under multi-version concurrency control
requires checking for update conflicts. All told, the reliance on a
globally agreed total order necessitates a minimum of one round-trip
to a designated master or coordination service for each of these
classic algorithms.  The cost of round trips will be determined by the
deployment environment, as we saw in Section~\ref{sec:motivation}; we
will demonstrate this cost on public cloud infrastructure in
Section~\ref{sec:prototype}.

\vspace{.5em}\noindent\textbf{Non-serializability} Most existing
distributed implementations of weak isolation are not highly
available. Lock-based proposals such as those used to provide weak
isolation in Gray's original proposal~\cite{gray-isolation} do not
degrade gracefully in the presence of partial failures. (Note,
however, that lock-based protocols \textit{do} offer the benefit of
recency guarantees.) While multi-versioned storage systems allow for a
variety of transactional guarantees, few offer traditional weak
isolation (e.g., non-``tentative update'' schemes) in this context.
The MDCC~\cite{mdcc} protocol offers Read Committed isolation with
Lost Update avoidance but is similarly unavailable due to its reliance
on preventing write conflicts. Chan and Gray's read-only transactions
have item-cut isolation with causal consistency and TA (session
\textit{PL-2L}~\cite{adya}) but are unavailable in the presence of coordinator
failure and assume serializable update transactions~\cite{readonly};
this is similar to read-only and write-only transactions more recently
proposed by Eiger~\cite{eiger}. Causal Serializability offers a
similar model (with unavailable implementation): causal consistency
with a variant of Read Uncommitted between transactions that write to
the same data item~\cite{raynal-causal}.  Brantner's S3
database~\cite{kraska-s3} and Bayou~\cite{sessionguarantees} can all
provide variants of session \textit{PL-2L} with high availability, but none
provide this HAT functionality without substantial
modification. Swift~\cite{swift} and bolt-on causal
consistency~\cite{bolton} are closest to providing maximum sticky HAT
semantics. As we have seen, it is possible to implement many
guarantees weaker than serializability---including guarantees that are
achievable with high availability---and still not achieve high
availability.

@@

\vspace{.5em}\noindent\textbf{Commutativity and Monotonicity} As
evidenced by the CALM Theorem~\cite{calm} and by Commutative and
Replicated Data Types~\cite{crdt}, if updates logically commute, then
they can often be safely performed in different orders at different
replicas. Accordingly, as long as all writes are delivered to all
replicas, then a system executing monotonic logic may not suffer from
application-level consistency anomalies as a result of Lost Update or
Write Skew anomalies. However, applications with non-monotonic state
mutation will not, in general, be able to maintain application-level
consistency constraints with HATs alone---in particular, applications
requiring bounded update visibility latency should opt for
unavailability.

@@

To further understand the performance implications of HAT guarantees
in a real-world environment, we implemented a HAT database
prototype. Across a range of deployments on public cloud
infrastructures, we found that, as Section~\ref{sec:latency}'s
measurements suggested, ``strongly consistent'' algorithms incur
substantial latency penalties: over WAN, 10 to 100 times higher than
their HAT counterparts. In light of recent algorithms for providing
causal consistency~\cite{bolton, cops, swift}, we focus our HAT
analysis on TA (algorithm from Section~\ref{sec:ta}, described in
greater in our Technical Report~\cite{hat-tr}), which we believe to be
the most novel of our algorithms. We find that HAT properties---while
more resource-intensive than basic eventual consistency---scale well
to many servers.

\vspace{.5em}\noindent\textit{Implementation.} Our prototype database
is a partially replicated (hash-based, partitioned) key-value backed
by LevelDB and implemented in Java using Apache Thrift. It currently
supports eventual consistency (hereafter, \texttt{eventual};
last-writer-wins RU with standard all-to-all anti-entropy between
replicas) and HAT RC and TA with RC guarantees (hereafter, \texttt{TA}). It also supports
non-HAT operation whereby all operations for a given key are routed to
a (randomly) designated \texttt{master} replica for each key
(hereafter, \texttt{master}; i.e., PNUTS~\cite{pnuts}) as well as
distributed two-phase locking. Servers are durable: they synchronously
write to LevelDB before responding to client requests, while new
writes in TA are sent to a write-ahead log, which is synchronously
flushed to disk.

\vspace{.5em}\noindent\textit{Configuration.} We deploy the database
in \textit{clusters}---disjoint sets of database servers that each
contain a single, fully replicated copy of the data---typically across
datacenters, sticking all clients within a datacenter to their
respective cluster (trivially providing read-your-writes and monotonic
reads guarantees). By default, we use 5 Amazon EC2 \texttt{m1.xlarge}
instances as servers in each cluster. For our workload, we link our
client library to the YCSB benchmark~\cite{ycsb}, which is well-suited
to LevelDB's key-value schema, grouping every eight YCSB operations
from the default workload (50\% reads, 50\% writes) to form a
transaction. We increase the number of keys in the workload from the
default 1,000 to 100,000 with uniform random key access, keeping
the default value size of $1KB$, and running YCSB for 180 seconds per
configuration.

@@

\vspace{.5em}\noindent\textit{Geo-replication.} We first deploy the
database prototype across an increasing number of
datacenters. Figure~\ref{fig:wan-exp}A shows that, when operating two
clusters within a single datacenter, mastering each data item results
in approximately half the throughput and double the latency of
\texttt{eventual}. This is because HAT models are able to utilize
replicas in both clusters instead of always contacting the (single)
master. \texttt{RC}---essentially \texttt{eventual} with
buffering---is almost identical to \texttt{eventual}, while
\texttt{TA}---which incurs two writes for every client-side write
(i.e., new writes are sent to the WAL then subsequently moved into
LevelDB once pending stable)---achieves ~75\% of the throughput. Latency
increases linearly with the number of YCSB clients due to contention
within LevelDB.

In contrast, when the two clusters are deployed across the continental
United States (Figure~\ref{fig:wan-exp}B), the average latency of
\texttt{master} increases to $300$ms (a $278$--$4257\%$ latency
increase; average $37$ms latency per operation). For the same number
of YCSB client threads, \texttt{master} has substantially lower
throughput than the HAT configurations. Increasing the number of YCSB
clients \textit{does} increase the throughput of \texttt{master}, but
our Thrift-based server-side connection processing did not gracefully
handle more than several thousand concurrent connections. In contrast,
across two datacenters, the performance of eventual, RC, and TA are
near-identical to a single-datacenter deployment.

When five clusters (as opposed to two, as before) are deployed across
the five EC2 datacenters with lowest communication cost
(Figure~\ref{fig:wan-exp}C), the trend continues: \texttt{master}
latency increases to nearly $800$ms per transaction. As an attempt at
reducing this overhead, we implemented and benchmarked a variant of
quorum-based replication as in Dynamo~\cite{dynamo}, where clients
sent requests to all replicas, which completed as soon as a majority
of servers responded (guaranteeing regular
semantics~\cite{herlihy-art}); this strategy (not pictured) did not
substantially improve performance due to the network topology and
because worst-case server load was unaffected. With five clusters,
TA's relative throughput decreased: every YCSB \texttt{put} operation
resulted in four \texttt{put} operations on remote replicas and,
accordingly, the cost of anti-entropy increased (e.g., each server
processed four replica's anti-entropy as opposed to one before,
reducing the opportunity for batching and decreasing available
resources for incoming client requests). This in turn increased
garbage collection activity and, more importantly, IOPS when compared
to Eventual and RC, causing TA throughput to peak at around half of
Eventual. With in-memory persistence (i.e., no LevelDB or WAL),
\texttt{TA} throughput was within 20\% of \texttt{eventual}.

We have intentionally omitted performance data for two-phase
locking. \texttt{master} performed \textit{far} better than our
textbook implementation, which, in addition to requiring a WAN
round-trip per operation, also incurred substantial overheads due to
mutual exclusion via locking. We expect that, while techniques like
those recently proposed in Calvin~\cite{calvin} can reduce the
overhead of serializable transactions by avoiding locking, our
mastered implementation and the data from Section~\ref{sec:latency}
are reasonable lower bounds on latency.

@@

\vspace{.5em}\noindent\textit{Transaction length.} As shown in
Figure~\ref{fig:txlen} (clusters in Virginia and Oregon), throughput
of \texttt{eventual}, \texttt{RC}, and \texttt{master} operation are
unaffected by transaction length. In contrast, \texttt{TA} throughput
decreases linearly with increased transaction length: with $1$
operation per transaction, \texttt{TA} throughput is within 18\% of
\texttt{eventual} ($34$ bytes overhead), and with $128$ operations per
transaction, \texttt{TA} throughput is within $60\%$ ($1898$ bytes
overhead). This reflects our TA algorithm's metadata requirements,
which are proportional to transaction length and consume IOPS and
network bandwidth.

\vspace{.5em}\noindent\textit{Read proportion.} Our default (equal)
proportion of reads and writes is fairly pessimistic: for example,
Facebook reports $99.8\%$ reads for their workload~\cite{eiger}. As
shown in Figure~\ref{fig:rprop} (clusters in Virginia and Oregon),
with all reads, \texttt{TA} is within $4.8\%$ of \texttt{eventual};
with all writes, \texttt{TA} is within $33\%$, and the throughput of
\texttt{eventual} decreases by $288.8\%$ compared to all reads. At
$99.8\%$ reads, \texttt{TA} incurs a $7\%$ overhead ($5.8\%$ for in-memory
storage).

\vspace{.5em}\noindent\textit{Scale-out.} One of the key benefits of
our HAT algorithms is that they are shared-nothing, meaning they
should not compromise scalability. Figure~\ref{fig:scaleout} shows
that varying the number of servers across two clusters in Virginia and
Oregon (with $15$ YCSB clients per server) results in linear scale-out
for \texttt{eventual}, \texttt{RC}, and \texttt{TA}. \texttt{RC} and
\texttt{eventual} scale perfectly with an approximately $5$x
throughput increase moving from $5$ to $25$ servers per cluster. For
the same configuration, \texttt{TA} scales by $3.8$x, achieving over
$260,000$ operations per second. \texttt{TA} performance is due to
resource contention---with a memory-backed database (instead of
LevelDB), \texttt{TA} scales by $4.25$x (not shown)---and TA-related
performance heterogeneity across servers (Calvin's authors report
similar heterogeneity on EC2~\cite{calvin}). Nevertheless, this linear
scalability is not typical of traditional database system
designs~\cite{gray-isolation}.

\vspace{.5em}\noindent\textit{Summary.} Our experimental prototype
confirms our earlier analytical intuitions. HAT systems can provide
useful semantics without substantial performance penalties. In
particular, our TA algorithm can achieve throughput competitive with
eventual consistency at the expense of increased disk and network
utilization. Perhaps more importantly, all HAT algorithms circumvent
high WAN latencies inevitable with non-HAT implementations. Our results
highlight Deutsch's observation that ignoring factors such as latency
can ``cause big trouble and painful learning
experiences''~\cite{fallacies-deutsch}---in a single-site context,
paying the cost of coordination may be tenable, but, especially as
services are geo-replicated, costs increase.
